---
title: "R on FHIR"
author: "Alexander V. Alekseyenko (alexander.alekseyenko@gmail.com)"
date: "`r date()`"
output: 
  html_document: 
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: true
---


# Preliminaries
This session was solicited by Doug Fridsma.

Taught by Grahame Grieve.

The workshop uses POSTMan (https://www.getpostman.com/downloads/)

The actual R exercises have been preceded with examples of querying mimic data using FHIR.

See hapifhir.io for a Hapi server.

Load all the libraries.
```{r}
library(httpuv); packageVersion("httpuv")
library(devtools); packageVersion("devtools")
# install_github("FirelyTeam/RonFHIR")
library(RonFHIR); packageVersion("RonFHIR")
library(stringr); packageVersion("stringr")
library(ggplot2); packageVersion("ggplot2")
```


Open client connection
```{r}
client <- fhirClient$new("http://mimic.fhir.org/r3")
```


Simple search
```{r}
bundle <- client$search("Patient", c("gender=female", "birthdate=gt2082-06-27"))
```

What did we get back?
```{r}
class(bundle)
names(bundle)
summary(bundle)
# Don't do this; you'll flood your screen
#str(bundle)
```

Summary of the number of records and matches

```{r}
paste("Number of matching patients =", bundle$total)
paste("Number of records =", dim(bundle$entry)[1])
```

Print the records gender and date of birth (my solution)
```{r}
paste(bundle$entry$resource$gender, 
      bundle$entry$resource$birthDate,
      unlist(lapply(bundle$entry$resource$extension, "[", 1, 2)),
      sep=",")
```

Logic behind my solution:
```{r}
class(bundle$entry)
names(bundle$entry)

class(bundle$entry$resource)
names(bundle$entry$resource)

class(bundle$entry$resource$gender)
class(bundle$entry$resource$birthDate)
```

We'll need a function to extract religion.
```{r}
getReligion <- function(extl) {
  for (io in extl) {
      for (i in 1:nrow(io)) {
        if (io[i, "url"] == "http://hl7.org/fhir/StructureDefinition/patient-religion") {
          v <- io[i, "valueCodeableConcept"];
          return (v$text);
        }
      }
  }
  return ("??")
}
```



Print the records gender and date of birth (Grahame's solution)
```{r}
for(row in 1:nrow(bundle$entry)) {
  res <-bundle$entry[row, "resource"]
  print(paste(res$gender, res$birthDate, 
              getReligion(res$extension),  sep = ","))
}
```

# Example iterating through pages of results

Specify a search with 30 records per page
```{r}
bundle <- client$search("Patient", c("_count=30"))
```


```{r}
class(bundle)
names(bundle)
class(client)
names(client)
```


```{r}
class(client$continue)
```


```{r}
while(!is.null(bundle)) {
  print(c("================================= Number of records=", 
          dim(bundle$entry)[1]))
  print(
    paste(bundle$entry$resource$gender,
          bundle$entry$resource$birthDate,
          unlist(lapply(bundle$entry$resource$extension, "[", 1, 2)),
          sep=","))
  
  # Go to the next page of the bundle using FHIRs paging mechanism
  bundle <- client$continue(bundle)
}
```


# using graphQL example 1
What are we doing here?
```{r}
res <- client$qraphQL('{PatientList(){
                      gender birthDate religion: 
                      extension(url :
                  "http://hl7.org/fhir/StructureDefinition/patient-religion") 
                  {value : valueCodeableConcept { text } } }}');
pl <- res$data$PatientList
```


```{r}
pl
```

Note that religion did not print. See the magic below to make it work.

# using graphQL Example 2
```{r}
res <- client$qraphQL('{PatientList(){
                      gender birthDate religion: 
                      extension(url : 
                  "http://hl7.org/fhir/StructureDefinition/patient-religion") 
                  @flatten {value : valueCodeableConcept 
                  @flatten { religion: text } } }}');
pl <- res$data$PatientList
```


easy now...
```{r}
pl
```


# Using OAuth

Need lots of information from the server:
* App-Name: (“Mimic” for Mimic Server)
* Client-Id:  (c.5 for Mimic server)
* Client-Secret: (cfe2cc3e-d4ca-49f4-8366-10064f4eda5c for Mimic server)
* Auth-Endpoint: https://mimic.fhir.org/r3/auth/auth
* Token-Endpoint:  https://mimic.fhir.org/r3/auth/token
* Username / Password: gg / Mimic

```{r eval=FALSE}
client <- fhirClient$new("https://mimic.fhir.org/r3");
app <- httr::oauth_app(appname = "Mimic", 
                       "c.5", 
                       "cfe2cc3e-d4ca-49f4-8366-10064f4eda5c")
scopes <- c("patient/*.read")
oauth_endpoint <- 
  httr::oauth_endpoint(authorize = 
                         paste("https://mimic.fhir.org/r3/auth/auth", 
                               "?aud=", "https://mimic.fhir.org/r3", 
                               "&state=", runif(1), sep=""), 
                       access = "https://mimic.fhir.org/r3/auth/token")
token <- httr::oauth2.0_token(endpoint = oauth_endpoint, 
                              app = app, scope = scopes)

client$setToken(token$credentials$access_token)
```



# O2 Saturation example
Here we will build historic $O_2$ saturation dataset for a patient.

## Utility functions
We'll need a couple of functions to do date format conversion.
```{r}
fixDateTime <- function (s) {
  if(nchar(s) == 10) {
    return (paste(s, "00:00:00"));
  } else {
    return (str_replace(s,"T", " "));
  }
}

fixDateTime2 <- function (s) {
  ifelse(nchar(s) == 10, paste(s, "00:00:00"), sub("T", " ", s))
}
```


```{r}
client <- fhirClient$new("http://mimic.fhir.org/r3")

bundle <- client$search("Observation", 
                        c("code=2708-6", 
                          "subject=Patient/30831", 
                          "date=lt2130-08-01"))
```

```{r}
dateList <- list();
valueList <- list();
index <- 1

while(!is.null(bundle)) {
  for(row in 1:nrow(bundle$entry)) {
    res <-bundle$entry[row, "resource"]
    dateList[index] = fixDateTime(res$effectiveDateTime);
    valueList[index] = res$valueQuantity$value;
    index = index+1;
  }
  bundle <- client$continue(bundle)
}

df <- do.call(rbind, 
              Map(data.frame, 
                  DATE=dateList, 
                  SAT=valueList))
df
df$DATET <- as.POSIXct(df$DATE,tz=Sys.timezone())
df
```

Now let's actually plot this
```{r}
ggplot(data = df, aes(x = DATET, y = SAT)) +
  geom_point() + geom_line() +
  labs(x = "Date",
       y = "O2 Sat",
       title = "Oxygen Saturation")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```


## The same using graphQL
```{r}
res <- client$qraphQL('{ObservationList(code: "2708-6",
                      subject: "Patient/30831", 
                      date: "lt2130-08-01") 
                      {date : effectiveDateTime 
                      valueQuantity @flatten { value } }}');
df <- res$data$ObservationList
df$DATET <- as.POSIXct(fixDateTime2(df$date), 
                       tz=Sys.timezone())
df
```


Plotting
```{r}
ggplot(data = df, aes(x = DATET, y = value)) +
  geom_point() + geom_line() +
  labs(x = "Date",
       y = "O2 Sat",
       title = "Oxygen Saturation")+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) 
```


# More on graphQL
see [https://graphql.org]


